Data Exploration: IUCN and AquaMaps

### 2015-11-17 data_explore.Rmd
adding plot of area ratio to overlap, and representative species maps

---
title: "Data Exploration: IUCN and AquaMaps"
output: html_document
---

@@

###get just the aquamaps cells
  am_spp_map <- spp_map %>%
    filter(!is.na(am_prob))%>%
    dplyr::select(loiczid, am_sid, am_prob) %>% #add in am_prob when we want to look at probability
    unique()

@@

   #iucn map cells
  cells_iucn<-iucn_map%>%
              filter(iucn_pres == TRUE)%>%
              .$loiczid
  cells_am <- am_spp_map%>%
                filter(!is.na(am_sid))%>%
                .$loiczid

@@

  spp_list[i, ]$sm_range <- ifelse(length(cells_am)<length(cells_iucn), 'AM', 'IUCN') #what dataset has the smaller range
  spp_list[i, ]$sm_perc <- length(cells_overlap)/min(length(cells_am), length(cells_iucn)) * 100 
    #this is the total percent of smaller range encompassed in the larger range

@@

plot(r_am_spp, col = 'orange', 
     main = "Yellow-speckled Chromis", 
     useRaster = FALSE, 
     add = TRUE, 
     axes = FALSE, 
     box = FALSE, 
     legend = FALSE)

@@

legend("topright", 
       legend = c("AquaMaps", "IUCN"), 
       fill = c('orange', 'mediumpurple'))

@@

hist_sm_within_lg <- ggplot(spp_df4, aes(sm_perc, fill = sm_range)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 4, alpha = .8) +
  scale_fill_discrete(name = "Dataset with smaller range", labels = c("AquaMaps", "IUCN")) + 
  labs(x = 'Percent of smaller range that falls within the larger range', 
       y = 'Proportion of all species (n = 1962)')

@@

@@

dir_aquamaps <- file.path(dir_N, "git-annex/globalprep/SPP_ICO/raw/aquamaps_2014")

spp_cells <- fread(input = file.path(dir_aquamaps, "tables/ohi_hcaf_species_native.csv"), verbose = FALSE, header = TRUE, 
                  colClasses = c('NULL', 'character', NA, 'numeric', 'NULL', 'NULL')) #1 min 49 sec

setnames(spp_cells, c('SpeciesID', 'CsquareCode', 'probability'), c("species_id", "csquare_code", "probability"))

@@

spp       <- fread(input = file.path(dir_aquamaps, "tables/ohi_speciesoccursum.csv"), 
                  verbose = FALSE, header = TRUE) %>%
              mutate(scientific = paste(Genus, Species, sep = " "))
spp <- as.data.frame(spp)

@@

* Baseline is: preferring IUCN rangemaps over Aquamaps rangemaps; Aquamaps "presence" threshold is set at 40% or greater.
* Scenario 1: preferring IUCN rangemaps over Aquamaps; AM presence threshold set at >= 1% (effectively, all occurrences).
* Scenario 2: preferring AM maps over IUCN; AM presence >= 40%.
* Scenario 3: preferring AM over IUCN; AM presence >= 1%.

>>>>

---
title: 'Data Exploration: IUCN and AquaMaps'
output: html_document
---

@@

###get just the aquamaps cells
  am_spp_map <- spp_map %>%
    filter(!is.na(am_prob))%>%
    dplyr::select(loiczid, am_sid, am_prob) %>% 
      ### add in am_prob when we want to look at probability
    unique()

@@

   #iucn map cells
  cells_iucn<-iucn_map %>%
    filter(iucn_pres == TRUE) %>%
    .$loiczid
  cells_am <- am_spp_map %>%
    filter(!is.na(am_sid)) %>%
    .$loiczid

@@

  spp_list[i, ]$sm_range <- ifelse(length(cells_am)<length(cells_iucn), 'AM', 'IUCN') 
    ### what dataset has the smaller range
  spp_list[i, ]$sm_perc <- length(cells_overlap)/min(length(cells_am), length(cells_iucn)) * 100 
    ### this is the total percent of smaller range encompassed in the larger range

@@

plot(r_am_spp, 
     col       = 'orange', 
     main      = 'Yellow-speckled Chromis', 
     useRaster = FALSE, 
     add       = TRUE, 
     axes      = FALSE, 
     box       = FALSE, 
     legend    = FALSE)

@@

legend('topright', 
       legend = c('AquaMaps', 'IUCN'), 
       fill   = c('orange', 'mediumpurple'))

@@

hist_sm_within_lg <- ggplot(spp_df4, aes(sm_perc, fill = sm_range)) + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 4, alpha = .8) +
  scale_fill_discrete(name = 'Dataset with smaller range', labels = c('AquaMaps', 'IUCN')) + 
  labs(x = 'Percent of smaller range that falls within the larger range', 
       y = 'Proportion of all species (n = 1962)')

@@

``` {r pct overlap vs ratio of area, echo = FALSE, warning = FALSE, message = FALSE}

spp_list <- spp_list %>%
  mutate(area_ratio = am_area / iucn_area,
         area_ratio = ifelse(area_ratio > 1, 1 / area_ratio, area_ratio),
         area_ratio = area_ratio * 100)

### Inspect each quadrant to find representative species
quadrant1_spp <- spp_list %>%
  filter(area_ratio > 90 & sm_perc > 90)
quadrant2_spp <- spp_list %>%
  filter(area_ratio > 25 & area_ratio < 30) %>%
  filter(sm_perc > 90)
quadrant4_spp <- spp_list %>%
  filter(sm_perc > 25 & sm_perc < 30) %>%
  filter(area_ratio > 90)
quadrant3_spp <- spp_list %>%
  filter(area_ratio < 25 & sm_perc < 25)

spp_quads <- c('Kajikia albida', 'Galaxea cryptoramosa', 'Acanthurus nigroris', 'Metanephrops australiensis', 'Conus magnificus')

spp_list_label <- spp_list %>%
  filter(sciname %in% spp_quads)

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(area_ratio, sm_perc, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  geom_point(data = spp_list_label, color = 'black', size = 4) + 
  geom_text(data = spp_list_label, 
            aes(label = sciname), 
            color = 'black',
            size = 4,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

ggsave('figures/AreaRatiovsPercOverlap.png', width = 6, height = 4)

print(scatter_arearatio_vs_overlap)

```{r plot_spp_from_quadrants, echo = FALSE, warning = FALSE, message = FALSE}

for (spp in spp_quads) { # spp = 'Kajikia albida'
  map <- get_spp_map(spp)

spp_pts <- as.data.frame(rasterToPoints(r_iucn_spp)) %>%
  full_join(as.data.frame(rasterToPoints(r_am_spp)),
            by = c('x', 'y')) %>%

spp_plot <- ggplot(spp_pts, aes(x = x, y = y)) +
  geom_raster(aes(fill = iucn_pres), alpha = .5) +
  geom_raster(aes(fill = am_prob), alpha = .5) + 
  borders('world', color='gray40', fill='gray45') + # create a layer of borders

### plot raster of species: IUCN map, AM map, world map, legend
plot(r_iucn_spp, 
     col       = '#ff6666', 
     main      = spp,
     alpha     = 1,
     useRaster = FALSE, 
     axes      = FALSE, 
     box       = FALSE, 
     legend    = FALSE)

plot(r_am_spp, 
     col       = '#6666ff', 
     alpha     = .4,
     useRaster = FALSE, 
     add       = TRUE, 
     axes      = FALSE, 
     box       = FALSE, 
     legend    = FALSE)

legend('topright', 
       legend = c('AquaMaps', 'IUCN'), 
       fill   = c('#6666ff', '#ff6666'))
}

@@

dir_aquamaps <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO/raw/aquamaps_2014')

spp_cells <- fread(input = file.path(dir_aquamaps, 'tables/ohi_hcaf_species_native.csv'), verbose = FALSE, header = TRUE, 
                  colClasses = c('NULL', 'character', NA, 'numeric', 'NULL', 'NULL')) #1 min 49 sec

setnames(spp_cells, c('SpeciesID', 'CsquareCode', 'probability'), c('species_id', 'csquare_code', 'probability'))

@@

spp       <- fread(input = file.path(dir_aquamaps, 'tables/ohi_speciesoccursum.csv'), 
                  verbose = FALSE, header = TRUE) %>%
              mutate(scientific = paste(Genus, Species, sep = ' '))
spp <- as.data.frame(spp)

@@

* Baseline is: preferring IUCN rangemaps over Aquamaps rangemaps; Aquamaps 'presence' threshold is set at 40% or greater.
* Scenario 1: preferring IUCN rangemaps over Aquamaps; AM presence threshold set at >= 1% (effectively, all occurrences).
* Scenario 2: preferring AM maps over IUCN; AM presence >= 40%.
* Scenario 3: preferring AM over IUCN; AM presence >= 1%.
